D3064R0: How to Avoid OOTA Without Really Trying

### 2024-03-21 ootavacuous.tex
Simplify inc.litmus and inc-range.litmus
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Create summary appendix and relegate a couple tests to it
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/inc-range.litmus @@
\caption{Only Atomic Increment, Optimized}
\label{lst:Only Atomic Increment, Optimized}
\end{listing}

Let us drop the single-thread-analysis constraint for
the moment, and suppose that the compiler is able to prove
that the only modification of a given atomic variable is to atomically
increment it, as is the case for \co{y} in
Listing~\ref{lst:Only Atomic Increment}.
Is it okay for a C++ compiler to transform this program into
Listing~\ref{lst:Only Atomic Increment, Optimized}?
An argument in favor is that the value of \co{y} must have
passed through the value 42 if \co{r0} and \co{r1}
bracket that value.
Arguments against might cite the added overhead of the invented load,
or the possibility that the check for the value 42 is intended for statistical
sampling.

@@

\subsection{Multi-Cycle Litmus Test}
\label{app:Multi-Cycle Litmus Test}

Listing~\ref{lst:Multi-Cycle Litmus Test}
expands
Listing~\ref{lst:Causality Test Case 4}
in a different way, moving to four threads but having only the
original two variables \co{x} and \co{y}.
The \co{herd7} tool shows multiple different possible OOTA cycles,
as expected.
However, the value returned by each thread's load directly affects the
value stored, so there is a semantic dependency in each thread.
Therefore, none of these OOTA cycles can arise in a compiler-based
C++ implementation.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-two-source.litmus @@
\caption{Multi-Cycle Litmus Test}
\label{lst:Multi-Cycle Litmus Test}
\end{listing}

\subsection{Non-LB Litmus Test}
\label{app:Non-LB Litmus Test}

Listing~\ref{lst:Non-LB Litmus Test}
shows a litmus test that can produce an OOTA cycle, but that is not
a member of the load-buffering (LB) class of litmus tests.
Suppose that lines~9 and~10 both somehow return the value 21.
Then line~12 will store the value 42 to \co{x}, which will cause
line~17 to store 21 to \co{y} and line~22 to store 21 to \co{z},
thus justifying the values initially loaded by lines~9 and~10.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-non-lb.litmus @@
\caption{Non-LB Litmus Test}
\label{lst:Non-LB Litmus Test}
\end{listing}

All thread's loads' return values affect the value stored, so all
threads have semantic dependencies between their respective relaxed
atomic loads and relaxed atomic stores.
Therefore, an OOTA cycle cannot form in a compiler-based C++
implementation.

Note that in a C++ implementation that (erroneously) permitted an OOTA
cycle to form, that cycle would involve both of \co{P0()}'s \co{r0}
and \co{r1} local variables.

>>>>

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/inc-range.litmus @@
\caption{Only Atomic Increment, Extended}
\label{lst:Only Atomic Increment, Extended}
\end{listing}

Let us drop the single-thread-analysis constraint for
the moment, and suppose that the compiler is able to prove
that the only modification of a given atomic variable is to atomically
increment it, as is the case for \co{y} in
Listing~\ref{lst:Only Atomic Increment}.
Is it okay for a C++ compiler to transform this program into
Listing~\ref{lst:Only Atomic Increment, Extended}?
An argument in favor is that the value of \co{y} must have
passed through the value 42 if \co{r0} and \co{r1}
bracket that value.
Arguments against might cite the added overhead of the invented load,
or the possibility that the check for the value 42 is intended for statistical
sampling.

@@

\subsection{Additional Litmus Tests}
\label{app:Additional Litmus Tests}

This section lists a couple of other litmus tests of interest:

\begin{itemize}
\item	The litmus test at
	\url{https://github.com/paulmckrcu/oota/blob/master/litmus/oota-two-source.litmus}
	shows multiple overlapping OOTA cycles.
\item	The litmus test at
	\url{https://github.com/paulmckrcu/oota/blob/master/litmus/oota-non-lb.litmus}
	shows that OOTA cycles are not confined to the load-buffering (LB)
	pattern.
\end{itemize}
